Non-word repetition in children learning Yélî Dnye

### 2021-10-18 manuscript.tex
fix nasaliz. diacritics in Table 1 (NB: new input file)

\abstract{
In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
}

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000). Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. This allows us to contribute data at the intersection of language typology, language acquisition, and individual variation, as presented in more detail below.

@@

NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007; \textbf{chiat2015non?}). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). NWR tasks probably tap many skills (for relevant discussion see Coady \& Evans, 2008; Santos, Frau, Labrevoit, \& Zebib, 2020). Non-words can be designed to try to isolate more narrowly certain skills; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015). Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

@@

The first research area is at the intersection of typology and phonological development. There has been an interest in adapting NWR to different languages, in part for applied purposes. In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, (\textbf{chiat2015non?}) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody. As a result, any one item created will be relatively easier if it more closely resembles real words in the language, making it difficult to balance difficulty when comparing children learning different languages. This previous literature also suggests some dimensions of difficulty -- an issue to which we return in the next subsection.

@@

In the present study, we thought it was relevant to represent the rich phonological inventory found in Yélî Dnye, by including a variety of phonetic targets. Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets. Phonologists and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of acquisition, focusing largely on phonotactics (Moreton \& Pater, 2012). The correlation between typological frequency and ease of acquisition is typically assumed to emerge from the following causal paths (which are not mutually exclusive):

@@

It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture -- a hypothesis discussed for instance in (\textbf{chiat2015non?}) (pp.~7-8; see also p.~5). In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer. A non-systematic meta-analyses does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.

@@

In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991). Therefore, it may not be a general index of, for instance, phonological development, but reflect only certain non-universal language skills. Indeed, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. We return to the question of what was measured here in the Discussion.

Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).\footnote{Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants. In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment. To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge. For instance, Tuller et al. (2018) employed a set of non-words judged to be language independent and two others that were more aligned with either French or German. The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language (``language-dependent'') than when they are ``language-independent.'' To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience. Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR (Farabolini, Taboh, Ceravolo, \& Guerra, 2021). In any case, this line of research focuses on links between exposure to a given language and NWR performance. In contrast, when we discuss linguistic or cultural differences, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}

@@

Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun. For example, Peute and colleagues' (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones. We hope the present study contributes to this growing line of work.

Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based. Some children in our sample will have started school. Reading and writing instruction is currently done only in English (other than writing one's name). This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school. This change from Yélî-initial instruction to English only occurred about 15 years ago. It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.

@@

The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some billeting others, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area. These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline. Land attribution for building one's home is decided collectively based on land availability.

Most Yélî parents are swidden horticulturalists, who occasionally fish. Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020). Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2020). Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to billet their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

@@

We add some observations that will help us integrate this study into the broader investigation of NWR across cultures. As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020). However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for. The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is also relatively infrequent in this community (Casillas, Brown, \& Levinson, 2020). Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2020). Thus, if input quantities in early childhood are a major determinant of NWR scores, we should observe similarly low NWR scores as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).

@@

This study was approved as part of a larger research effort by the second author. The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent. As discussed in subsection ``The Yélî community,'' the combination of collective child guardianship practices and common billeting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, in press). Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver. Regardless of whether they completed the task, children were given a small snack as compensation. Children who showed initial interest but then decided not to participate were also given the snack.

@@

\caption{\label{tab:tabstims}NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.}
\centering
\begin{tabular}[t]{llllllllll}
\toprule
\multicolumn{2}{c}{Practice} & \multicolumn{2}{c}{Monosyll} & \multicolumn{2}{c}{Bisyll} & \multicolumn{2}{c}{Trisyll} & \multicolumn{2}{c}{Tetrasyll} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10}
Orth. & Phon. & Orth. & Phon. & Orth. & Phon. & Orth. & Phon. & Orth. & Phon.\\
\midrule
nopimade & nɔpimæṭɛ & dp:a & ṭp̃æ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
poni & pɔni & dpa & ṭpæ & kańi & kæni & diyeto & ṭijɛtɔ & ńomiwake & nɔmiwækɛ\\
wî & wɯ & dpâ & ṭpɑ & kipo & kipɔ & meyadi & mɛjæṭi & todiwuma & tɔṭiwumæ\\
 &  & dpê & ṭpə & ńoki & nɔki & mituye & mitujɛ & wadikeńo & wæṭikɛnɔ\\
 &  & dpéé & ṭpeː & ńomi & nɔmi & ńademo & næṭɛmɔ &  & \\
\addlinespace
 &  & dpi & ṭpi & piwa & piwæ & ńayeki & næjɛki &  & \\
 &  & dpu & ṭpu & towi & tɔwi & ńuyedi & nujɛṭi &  & \\
 &  & gh:ââ & ɣ̃ɑː & tupa & tupæ & pedumi & pɛṭumi &  & \\
 &  & ghuu & ɣuː &  &  & tiwuńe & tiwunɛ &  & \\
 &  & kp:ââ & kp̃ɑː &  &  & tumowe & tumɔwɛ &  & \\
\addlinespace
 &  & kpu & kpu &  &  & widońe & wiṭɔnɛ &  & \\
 &  & lv:ê & lβʲ̃ə &  &  & wumipo & wumipɔ &  & \\
 &  & lva & lβʲæ &  &  &  &  &  & \\
 &  & lvi & lβʲi &  &  &  &  &  & \\
 &  & t:êê & t̃əː &  &  &  &  &  & \\
\addlinespace
 &  & tpê & tpə &  &  &  &  &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

Additionally, we estimated frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2020). That corpus was constituted by sampling from audio-recordings (7-9 hours long), collected as 10 children aged between 1 month and 3 years went about their day. The researchers selected 9 2.5-minute clips randomly and 11 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity). These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the people talking personally. For more details, please refer to Casillas, Brown, and Levinson (2020).

For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space. We then removed all English and Pidgin words. The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types. To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, sounds from our non-word items that did not occur in the corpus were not considered (i.e., they were declared NA so that they do not count for analyses).

@@

The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020), then ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to .wav clips. The assistant then listened through all these paired target-repetition clips randomized across children and repetitions, grouped such that all the clips of the same target were listened to in succession. For each clip, the assistant indicated in a notebook whether the child production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English. The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets. After completing all of her annotations, she double-checked them by listening to them twice.

@@

Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for younger children. Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.

@@

We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data. Insertions and deletions were very rare (metathesis was not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets). We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones). To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?

@@

The range of performance we observed overlapped with previously observed levels of performance. Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli. Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane'.

>>>>

\abstract{
In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N = 40, aged 3--10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations, and the extent to which they shed light on the factors universally structuring variation in phonological development at a global and individual level.
}

@@

Children's perception and production of phonetic and phonological units continues developing well beyond the first year of life, even extending into middle childhood (e.g., Hazan \& Barrett, 2000; Rumsey, 2017). Much of the evidence for later phonological development comes from non-word repetition (NWR) tasks. In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG), which has a large and unusually dense phonological inventory. This allows us to contribute data at the intersection of language typology, language acquisition, and individual variation, as presented in more detail below.

@@

NWR has been used to seek answers to a variety of theoretical questions, including what the links between phonology, working memory, and the lexicon are (Bowey, 2001), and how extensively phonological constraints found in the lexicon affect online production (Gallagher, 2014). NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders (Estes, Evans, \& Else-Quest, 2007; \textbf{chiat2015non?}). Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe in the context of diagnosing language impairments among bilingual children (Armon-Lotem, Jong, \& Meir, 2015; Chiat, 2015; COST Action, 2009; Meir, Walters, \& Armon-Lotem, 2016). NWR tasks probably tap into many skills (for relevant discussion see Coady \& Evans, 2008; Santos, Frau, Labrevoit, \& Zebib, 2020). Non-words can be designed to try to isolate certain skills more narrowly; for instance, one can choose non-words that contain real morphemes in order to load more on prior language experience, or non-words that are shorter to avoid loading on working memory (see a discussion in Chiat, 2015). Broadly, however, NWR scores will necessarily reflect to a certain extent phonological knowledge (to perceive the item precisely despite not having heard it before) as well as online phonological working memory (to encode the item in the interval between hearing it and saying it back) and flexible production patterns (to produce the item precisely despite not having pronounced it before).

@@

The first research area is at the intersection of typology and phonological development. There has been an interest in adapting NWR to different languages, in part for applied purposes. In a review of NWR as a potential task to diagnose language impairments among bilingual children in Europe, (\textbf{chiat2015non?}) discusses the impossibility of creating language-universal non-word items: Languages vary in their phonological inventory, sound sequencing (phonotactics), syllable structure, and word-level prosody. As a result, any one item created will be relatively easier if it more closely resembles real words in a language, making it difficult to balance difficulty when comparing children learning different languages. This previous literature also suggests some dimensions of difficulty---an issue to which we return in the next subsection.

@@

In the present study, we thought it was relevant to represent the rich phonological inventory found in Yélî Dnye, by including a variety of phonetic targets. Some of them are cross-linguistically rare, in that they are less common across languages than other sounds or phonetic targets. Phonologists, phoneticians, and psycholinguists have discussed the extent to which cross-linguistic frequency may reflect ease of processing and acquisition via diachronic language change. These works focus largely on phonotactics (Moreton \& Pater, 2012) perceptual parsing of the (ambiguous) linguistic signal (\textbf{beddor2009coarticulatory?}; \textbf{ohala1981listener?}), and individual differences in processing styles (\textbf{bermudezotero2015amphichronic?}); small but significant effects that may cumulatively drive language change via phonologization (see \textbf{yu2021toward?} for a recent review). Thus, the correlation between typological frequency and ease of acquisition is typically assumed to emerge from one or more of the following causal paths:

@@

It is possible that differences are due to language-specific characteristics, including the most common length of words in the lexicon and/or in child-experienced speech in that culture---a hypothesis discussed for instance in (\textbf{chiat2015non?}) (pp.~7-8; see also p.~5). In broad terms, one may expect languages with a lexicon that is heavily biased towards monosyllables to show greater length effects than languages where words tend to be longer. A non-systematic meta-analysis does not provide overwhelming support for this hypothesis {[}Cristia and Casillas (2021); SM1{]}.

@@

In view of these results, it is important to bear in mind that NWR is a task developed in countries where literacy is widespread, and it is considered an excellent predictor of reading; for instance, better than rhyme awareness (e.g., Gathercole, Willis, \& Baddeley, 1991). Therefore, it may not be a general index of phonological development, but instead reflect certain non-universal language skills. Indeed, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) present their task as being a good index of the development of ``short-hand-like'' representations specifically, which could thus miss, for example, more holistic phonological and phonetic representations. We return to the question of what was measured here in the Discussion.

Aside from Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020)'s hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).\footnote{Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants. In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment. To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge. For instance, Tuller et al. (2018) employed a set of non-words judged to be language independent and two others that were more aligned with either French or German. The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language (``language-dependent'') than when they are ``language-independent.'' To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience. Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR (Farabolini, Taboh, Ceravolo, \& Guerra, 2021). In any case, this line of research focuses on links between exposure to a given language and NWR performance. In contrast, when we discuss linguistic or cultural differences here, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).}

@@

Our knowledge of Yélî language development is growing (e.g., Brown, 2011, 2014; Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020; Liszkowski, Brown, Callaghan, Takada, \& de Vos, 2012), but research into Yélî phonological development has only just begun. For example, Peute and Casillas (In preparation) find that Yélî Dnye-learning children's early spontaneous consonant productions appear to exclusively feature simplex and typologically frequent phones. Other ongoing work on Yélî Dnye includes experiment-based infant phoneme discrimination data and errors made in elicited and spontaneous speech from young children, but these data are neither finalized nor yet externally reviewed (see Hellwig, Sarvasy, \& Casillas, provisionally accepted for more information). These data will help better inform our current analyses based on NWR in the future (e.g., regarding common sound substitutions) but are not critical for testing our current question about the general correlation between cross-linguistic phone frequency and NWR performance.

Before closing this section, it bears mentioning that the language has an established orthography, which includes distinct graphemes for all the contrasts on which our items are based. Some children in our sample will have started school. Reading and writing instruction is currently done only in English (other than writing one's name). This was probably not the case for the majority of mothers of the children in our sample, who will have learned to read and write in Yélî Dnye during their first three years at school. It is possible that there is also some home teaching of Yélî reading and writing, notably for reading the bible.

@@

The typical household in our dataset includes seven individuals (typically, a mixed-sex couple and children---their own and possibly some others staying with them, as discussed in the next paragraph) and is situated among a collection of four or more other households, with structures often arranged around an open grassy area. These household clusters are organized by patrilocal relation, such that they typically comprise a set of brothers, their wives and children, and their mother and father, with neighboring hamlets also typically related through the patriline. Land attribution for building one's home is decided collectively based on land availability.

Most Yélî parents are swidden horticulturalists, who occasionally fish. Within a group of households, it is often the case that older adolescents and adults spend their day tending to their farm plots (which may not be nearby), bringing up water from the river, washing clothes, preparing food, and engaging in other such activities. Starting around age two years, children more often spend large swaths of their day playing, swimming, and foraging for fruit, nuts, and shellfish in large (\textasciitilde10 members) independent and mixed-age child play groups (Brown \& Casillas, in press; Casillas, Brown, \& Levinson, 2020). Formal education is a priority for Yélî families, and many young parents have themselves pursued additional education beyond what is locally available (Casillas, Brown, \& Levinson, 2020). Local schools are well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day), so it is very common for households situated close to a school to host their school-aged relatives during the weekdays for long segments of the school year. Children start school often at around age seven, although the precise age depends on the child's readiness, as judged by their teacher.

@@

We add some observations that will help us integrate this study into the broader investigation of NWR across cultures. As mentioned previously, there is one report of relatively low NWR scores among the Tsimane', which the authors of that paper interpret as consistent with long-term effects of low levels of infant-directed speech (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020). However, Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020) also point out that this is based on between-paper comparisons, and thus methods and myriad other factors have not been controlled for. The Yélî community can help us gain new insights into this matter because direct speech to children under 3;0 is comparably infrequent in this community (in fact it may be infrequent in many settings, including urban ones \textbf{bunceURcross?}), and additionally shares other societal characteristics wih the Tsimane' {[}e.g., is rural and relies on farming, children grow up in wide familial networks, etc; Casillas, Brown, and Levinson (2020){]}. Although infant-directed speech has been measured in different ways among the Tsimane' and the Yélî communities, our most comparable estimates at present suggest that Tsimane' young children are spoken to about 4.2 minutes per hour (Scaff, Stieglitz, Casillas, \& Cristia, 2021), and Yélî children about 3.6 minutes per hour (Casillas, Brown, \& Levinson, 2020). Thus, if these input quantities in early childhood relate to lower NWR scores later in life, we should observe similarly low NWR scores here as in Cristia, Farabolini, Scaff, Havron, and Stieglitz (2020).

@@

This study was approved as part of a larger research effort by the second author. The line of research was evaluated by the Radboud University Faculty of Social Sciences Ethics Committee (Ethiek Commissie van de faculteit der Sociale Wetenschappen; ECSW) in Nijmegen, The Netherlands (original request: ECSW2017-3001-474 Manko-Rowland; amendment: ECSW-2018-041), including the use of verbal (not written) consent. As discussed in subsection ``The Yélî community,'' the combination of collective child guardianship practices and common hosting of school-aged children for them to attend school is that adult consent often comes from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. Child assent is also culturally pertinent, as independence is encouraged and respected from toddlerhood (Brown \& Casillas, in press). Participation was voluntary; children were invited to participate following indication of approval from an adult caregiver. Regardless of whether they completed the task, children were given a small snack as compensation. Children who showed initial interest but then decided not to participate were also given the snack.

@@

\caption{\label{tab:tabstims}NWR stimuli in orthographic (Orth.) and phonological (Phon.) representations.}
\centering
\begin{tabular}[t]{llllllllll}
\toprule
\multicolumn{2}{c}{Practice} & \multicolumn{2}{c}{Monosyll} & \multicolumn{2}{c}{Bisyll} & \multicolumn{2}{c}{Trisyll} & \multicolumn{2}{c}{Tetrasyll} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6} \cmidrule(l{3pt}r{3pt}){7-8} \cmidrule(l{3pt}r{3pt}){9-10}
Orth. & Phon. & Orth. & Phon. & Orth. & Phon. & Orth. & Phon. & Orth. & Phon.\\
\midrule
nopimade & nɔpimæṭɛ & dp:a & ṭpæ̃ & kamo & kæmɔ & dimope & ṭimɔpɛ & dipońate & ṭipɔnætɛ\\
poni & pɔni & dpa & ṭpæ & kańi & kæni & diyeto & ṭijɛtɔ & ńomiwake & nɔmiwækɛ\\
wî & wɯ & dpâ & ṭpɑ & kipo & kipɔ & meyadi & mɛjæṭi & todiwuma & tɔṭiwumæ\\
 &  & dpê & ṭpə & ńoki & nɔki & mituye & mitujɛ & wadikeńo & wæṭikɛnɔ\\
 &  & dpéé & ṭpeː & ńomi & nɔmi & ńademo & næṭɛmɔ &  & \\
\addlinespace
 &  & dpi & ṭpi & piwa & piwæ & ńayeki & næjɛki &  & \\
 &  & dpu & ṭpu & towi & tɔwi & ńuyedi & nujɛṭi &  & \\
 &  & gh:ââ & ɣɑ̃ː & tupa & tupæ & pedumi & pɛṭumi &  & \\
 &  & ghuu & ɣuː &  &  & tiwuńe & tiwunɛ &  & \\
 &  & kp:ââ & kpɑ̃ː &  &  & tumowe & tumɔwɛ &  & \\
\addlinespace
 &  & kpu & kpu &  &  & widońe & wiṭɔnɛ &  & \\
 &  & lv:ê & lβʲə̃ &  &  & wumipo & wumipɔ &  & \\
 &  & lva & lβʲæ &  &  &  &  &  & \\
 &  & lvi & lβʲi &  &  &  &  &  & \\
 &  & t:êê & tə̃ː &  &  &  &  &  & \\
\addlinespace
 &  & tpê & tpə &  &  &  &  &  & \\
\bottomrule
\end{tabular}
\end{table}

@@

Additionally, we estimated the usage frequency of the phones present in the target items in a corpus of child-centered recordings (Casillas, Brown, \& Levinson, 2020). That corpus was constituted by sampling from audio-recordings (7--9 hours long), collected as 10 children aged between 1 month and 3 years went about their day. The researchers selected 9 2.5-minute clips randomly and 11 1- or 5-minute clips by hand (selected to represent peak turn-taking and child vocal activity). These clips were segmented and transcribed by the lead researcher and a highly knowledgeable local assistant, who speaks Yélî Dnye natively, has ample experience in this kind of research, and often knew all the recorded people personally. For more details, please refer to Casillas, Brown, and Levinson (2020).

For the present study, we extracted the transcriptions of adult speech (i.e., removing key child and other children's speech) and split them into words using white space. We then removed all English and Tok Pisin words. The resulting corpus contained a total of 18,934 word tokens of 1,686 unique word types. To get our phone frequency measure, we counted the number of word types in which the phone occurred, and applied the natural logarithm.\footnote{We also carried out analyses using token (rather than type) phone frequency, but this measure was not correlated with whole-item NWR scores, and therefore the fact that it did not explain away the predictive value of cross-linguistic phone frequency was less informative than the relationship discussed in the Results section.} Here, unattested sounds were not considered (i.e., they were declared NA so that they do not count for analyses). Note that the resulting values estimate usage frequencies for very young children's input and, while this is somewhat different from what our older participants experience on a daily basis, we can expect that this is a reasonable approximation of the early input that formed the foundation of their phonological knowledge.

@@

The first author then annotated the onset and offset of all children's productions from the audio recording using Praat audio annotation software (Boersma \& Weenink, 2020), then ran a script to extract these tokens, pairing them with their original auditory target stimulus, and writing these audio pairs out to .wav clips. The assistant then listened through all these paired target-repetition clips randomized across children and repetitions, grouped such that all the clips of the same target were listened to in succession. For each clip, the assistant indicated in a notebook whether the child production was a correct or incorrect repetition and orthographically transcribed the production, noting when the child uttered a recognizable word or phrase and adding the translation equivalent of that word/phrase into English. The assistant was also provided with some general examples of the types of errors children made without making specific reference to Yélî sounds or the items in the elicitation sets. Because the phonological inventory is so acoustically packed and because annotation was done based on audio data alone, it might be easy to mis-identify a segment, the assistant double-checked all of her annotations by listening to them and assessing them a second time, once she had completed a full first round.

@@

{\centering \includegraphics[width=0.65\linewidth]{nwr.by.freq.ITEM}

\caption{NWR scores for individual target words as a function of the average frequency with which each phone is found across languages.}\label{fig:fig-xling-freq}
\end{figure}

\begin{verbatim}
## pdf 
##   2
\end{verbatim}

Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = 0.22, SE ß = 0.09, p = 0.01): The effect of frequency was larger for younger children. Inspection of Figure \ref{fig:fig-xling-freq-age3} suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically uncommon phones.

@@

We addressed our first research question in a second way, by investigating patterns of error, looking at all attempts and not excluding errors resulting in real words, so as to base our generalizations on more data. Insertions and deletions were very rare (metathesis was not attested): there were only 17 instances of deleted vowels (\textasciitilde0.35\% of all vowel targets), and 13 instances of deleted consonants (\textasciitilde0.50\% of all consonant targets). We therefore focus our qualitative description here on substitutions: There were 813 cases of substitutions, \textasciitilde{} of the phones found collapsing across all children and target words, so that substitutions constituted the majority of incorrect phones (\textasciitilde{} of unmatched phones). To inform our understanding of how cross-linguistic patterns may be reflected in NWR scores, we asked: Is it the case that cross-linguistically less common and/or more complex phones are more frequently mispronounced, and more frequently substituted by more common ones than vice versa?\footnote{Note that a full anonymized, but age-annotated table of errors is provided in the project repository for those interested in a finer-grained analysis than what is presented here: ALEX?{]}.}

@@

The range of performance we observed overlapped with previously observed levels of performance. Paired with our thorough training protocol, we had interpreted the NWR scores among Yélî Dnye learners as indicating that our adaptations to NWR for this context were successful, even given a number of non-standard changes to the training phase and to the design of the stimuli. Additionally, it seemed that Yélî children showed comparable performance to others tested on a similar task, despite the many linguistic, cultural, and socioeconomic differences between this and previously tested populations, unlike the case that had been reported for the Tsimane' (Cristia, Farabolini, Scaff, Havron, \& Stieglitz, 2020).
